Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2014-08-27 draft.tex
Address issue pointed out by one of Nico's postdocs
More work on Sec IV and a first attempt at writing a conclusion

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \left(M^{-1}\right)_{ij} \est{S_j}\,.
\end{equation}

We emphasize that this methods hinges on the assumption in \refeq{linear-assumption}; we thus suggest to explicitly
test the linearity. One way of testing involves repeating the previous procedure
with $S_j^\text{true} = 1/z$, $z > 0$. If the linearity assumption holds, one should find
\begin{equation}
    M_{ij}(z) = z \est{S_i}\,,
\end{equation}
for a fixed value of $j$.

@@

\fred{$N$ used for the symbol $K$ of section I and II}
Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the pull distribution of the observable.
        All pull distributions can successfully be fitted to a gaussian distribution. As an example, we
        show the pull distributions for the observables $S_5$ and $S_7$ for a SM-like $B\to K^*\ell^+\ell^-$
        decays in \reffig{pulls}.
    \item We study the absolute uncertainty $\sigma_k(N)$ with respect to the angular observable $S_k$
        as a function of the number of simulated events $N$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \sigma_k(N) = \frac{\sigma_k(1)}{\sqrt{N}}
        \end{equation}
        with $\sigma_k(1) = \order{1}$, regardless of the size of $S_k$. The latter can best be shown
        for the uncertainties of two exampular observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
    \item For all toy studies, we produce plots of their Skewness and their Kurtosis as functions of
        $N$. We find in all cases that the distributions agree.
    \item \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of $N$.}
    \item \dots
\end{itemize}
We find that all of the above results show consistenly convergence of the distribution for the angular observables
towards a multivariate Gaussian distribution.

@@

We used the simulated events and fitted a likelihood function to them.
\begin{itemize}
    \item For a small number of simulated events, \danny{$N=?$}, we observe a significant bias in the pull distributions.
    \item For a large number of simulated events, \danny{$N=?$}, the uncertainty of the angular observables is
        smaller compared to the results of the method of moments.
\end{itemize}
\todo{compare robustness to likelihood fit when fit doesn't
converge. Why exactly doesn't it converge? What if physics model
wrong, say some angular terms are missing.}

@@

In this letter \dots
\begin{enumerate}
    \item errors on the means are in very good approximation gaussian.
    \item correlation are small and faithfully model a multivariate gaussian. \fred{$\Rightarrow$ easy transfer of knowledge to theory fits}
\end{enumerate}

@@

The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).
\dots
\danny{Potential people to read the manuscript: Jochen Dingfelder (Belle-II, Bonn), Ulrik Egede (LHCb, Imperial London), Christoph Langenbruch (LHCb, Warwick), Michel De Cian (LHCb, Heidelberg), Adrian Bevan (ATLAS, Queen Mary)  }
D.v.D would like to thank Gudrun Hiller and Martin Jung for discussing the results of \cite{Das:2014sra} prior to publication.

>>>>

It remains to invert the relation \refeq{linear-assumption}, so that one can determine the unfolded
angular observables $\lbrace \wwhat{S_k}\rbrace$ as
\begin{equation}
    \wwhat{S_i} = \danny{N_\eps} \left(M^{-1}\right)_{ij} \est{S_j}\,.
\end{equation}
\danny{Here $N_\eps$ is a normalization factor based on the detector acceptance function $\eps$. It emerges
    since the product $P(x) \eps(x)$ is not a PDF; it is not normalized appropriately\footnote{We thank
    X.X\danny{Marcin: Please insert whathisname} for pointing this out.}. We can determine $N_\eps$ by
    imposing
    \begin{gather}
        1 \equiv \int \rmdx{x} P(x | \lbrace \wwhat{S}_k\rbrace) = N_\eps \sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j}\int \rmdx{x} f_k(x)\,,\\
        \Rightarrow N_\eps \equiv \frac{1}{\sum_{k,j} \left(M^{-1}\right)_{kj} \est{S_j} \int \rmdx{x} f_k(x)}\,.
    \end{gather}
}

We emphasize that this method hinges on the assumption in \refeq{linear-assumption}\danny{which is well motivated:
the PDF is a linear combination of the basis functions $f_i$, and the moments are therefore a linear
functional of said basis functions. Nevertheless,}
we thus suggest to explicitly
test the linearity. One way of testing involves repeating the previous procedure
with $S_j^\text{true} = 1/z$, $z > 0$. If the linearity assumption holds, one should find
\begin{equation}
    M_{ij}(z) = z \est{S_i}\,,
\end{equation}
for a fixed value of $j$.

\danny{In order to better understand the detector aceptance effects, we propose two steps. For the sake of
    clarity of our argument, we assume in the following that the PDF $P$ is expanded in products of
    appropriately chosen Legendre polynomials to some multi-index order $\vec{l}_P^{(0)}$. Similarly,
    moments and detector effects are nominally determined up to some order $\vec{l}_M^{(0)}$, with
    $\vec{l}_M^{(0)} \leq \vec{l}_P^{(0)}$. Here and in the following all comparisons are to be read
    for each compoenent.\\
    For the first step, we remind that the unfolding matrix $M$ depends implicitly on the physical
    PDF $P$. We therefore recommend variation of the order of the expansion of $P$ to $\vec{l}_P \geq \vec{l}_P^{(0)}$
    component-wise. This way one can estimate the mixing of PDF terms of order $\vec{l}_P$ into moments of order $\vec{l}_M^{(0)}$
    or lower.\\
    In a second step, we assume using the nominal expansion of the PDF up to order $\vec{l}_P^{(0)}$.
    With moments of order $\vec{l}_M \geq \vec{l}_M^{(0)}$ one can therefore find the impact of the
    acceptance effects on the mixing of moments of order $\vec{l}_M^{(0})$ and below
    into moments beyond the order of the PDF.
    Our recommendation is not illustrated at the hand of numerical examples, which would necessitate
    access to realistic detector models. The latter are not available to persons outside the
    respective collaborations. The proposed studies are therefore beyond the scope of this work.
}

@@

\fred{$N$ used for the symbol $K$ of section I and II}
Our finding can be summarized as follows:
\begin{itemize}
    \item In all studied cases we observed not a single bias in the pull distribution of the observable.
        All pull distributions can successfully be fitted to a gaussian distribution. As an example, we
        show the pull distributions for the observables $S_5$ and $S_7$ for a SM-like $B\to K^*\ell^+\ell^-$
        decays in \reffig{pulls}.
    \item We study the absolute uncertainty $\sigma_k(K)$ with respect to the angular observable $S_k$
        as a function of the number of simulated events $K$. As expected for a multivariate gaussian
        distribution, we find that the absolute uncertainty is well fitted by
        \begin{equation}
            \sigma_k(N) = \frac{\sigma_k(1)}{\sqrt{N}}
        \end{equation}
        with $\sigma_k(1) = \order{1}$, regardless of the absolute size of $S_k$. The latter can best be shown
        for the uncertainties of two exampular observables. Taking again $S_5$ ($\simeq 27\%$) and $S_7$ ($\simeq 2\%$)
        for SM-like $B\to K^*\ell^+\ell^-$ decays, we show the absolute uncertainty in \reffig{errors}.
    \item For all toy studies, we produce plots of their skewness and their kurtosis as functions of
        $N$. We find in all cases that the mean skewness and kurtosis converge towards zero.
        \todo{Plot Skewness and Kurtosis for $S_5$ and $S_7$ as functions of the number of simulated events $K$.}
\end{itemize}
We find that all of the above results show consistently that the joint distrbution of the angular observables
converges towards a multivariate Gaussian distribution. We propose to describe the resulting distribution as
a multivariate Gauss when skewness and kurtosis are small enough. The precise targets for these quantities
are subject to the details of the physical analysis, and therefore outside the scope of our work.

@@

{\textbf{[dvd: can the following be removed?]}
We used the simulated events and fitted a likelihood function to them.
\begin{itemize}
    \item For a small number of simulated events, \danny{$N=?$}, we observe a significant bias in the pull distributions.
    \item For a large number of simulated events, \danny{$N=?$}, the uncertainty of the angular observables is
        smaller compared to the results of the method of moments.
\end{itemize}
\todo{compare robustness to likelihood fit when fit doesn't
converge. Why exactly doesn't it converge? What if physics model
wrong, say some angular terms are missing.}}

@@

We have carried out a combined analytical and numerical study of the method of moments; a method for
the extraction of angular observales from the angular distribution of a generic multi-body process.
We have studied the performance of the method of moments using pseudo data derived from the SM
predctions for one uniangular decay ($B\to K\ell^+\ell^-$) and one triangular decay ($B\to K^*\ell^+\ell^-$).
From this, we find rapid convergence of the joint likelihood of the angular observables towards a multivariate Gaussian.
We draw the conclusion that this method exhibits several benefits in the determination of angular observables when
compared with a maximum-likelihood fit.\\

First, we find no bias in the determinations of the angular observables even for a small number of events.
However, this comes at the expense of a slight increase of the order of $10\%$ for the statistical uncertainty
on the mean.
Second, in the absence of acceptance effects, a determination using the method of moments is correct and unbiased
even when not all aspects of the physical distribution are modelled. This is explicitl shown for the case of
higher partial waveds in multibody final states.
Third, we develop a systematic method for the dertermination of detector acceptance effects that lead to
dilution and mixing of the angular observables. We present a way to calculate the necessary unfolding matrix,
which is only computationally feasible when using the method of moments.\\
Fourth, the fact that the joint distribution of the results is Gaussian allows for an easy transfer of correlation
information to theory analyses.
Last but not least, the resulting distribution arises without additional constraints necessary for the convergence
of some maximum-likelihood fits. Therefore the results can be directly combined with similarly-obtained results from
a different experiment, without the need to remove the constraints of measurements that stem from different experiments.\\

In conclusion, we argue that the method of moments is a competetive alternative to maximum-likelihood fits as long
as angular distributions are involved. We wish to raise the interesting prospect of extending this very method to
applications with non-angular orthogonal bases for the PDF.

@@

The work of D.v.D has been supported by the Bundesministerium f\"ur Bildung und Forschung (BMBF).
We would like to thank Ulrik Egede and Konstantinos Petridis for insightful discussions.
\danny{Potential people to read the manuscript: Jochen Dingfelder (Belle-II, Bonn), Ulrik Egede (LHCb, Imperial London), Christoph Langenbruch (LHCb, Warwick), Michel De Cian (LHCb, Heidelberg), Adrian Bevan (ATLAS, Queen Mary)  }
D.v.D would like to thank Gudrun Hiller and Martin Jung for discussing the results of \cite{Das:2014sra} prior to publication.
